PyPy - How to \underline{not} write Virtual Machines for \mbox{Dynamic Languages}

### 2007-09-01 talk/esug2007/talk.tex
The ESUG2007 talk, based on the Dyla'07 one - sorry, I forgot to update to the latest version - but anyway the talk was mostly only the demo, not the slides.

\title{How to \underline{not} write Virtual Machines for \mbox{Dynamic Languages}}

\author{Carl Friedrich Bolz and Armin Rigo}

@@

\date{3rd Workshop on Dynamic Languages and Applications, July 31 2007}

@@

  \begin{itemize}
  \item
    implementing dynamic languages \par(with a focus on complicated ones)
  \item
    in a context of limited resources \par(academic, open source, or
    domain-specific)
  \end{itemize}
  \pause
  \begin{block}{
    Our point}
    \begin{itemize}
    \item
      Do not write virtual machines ``by hand''
    \item
      Instead, write interpreters in high-level languages
    \item
      Meta-programming is your friend
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Benefits of implementing on top of OO VMs}
  \begin{itemize}
  \item
    higher level of implementation
  \item
    the VM supplies a GC and mostly a JIT
  \item
    better interoperability than what the C level provides
  \item
    some proponents believe that eventually one single VM should be enough
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      both Jython and IronPython integrate well with their host OO VM
    \item
      XXX what else?
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The problems of OO VMs}
  \begin{itemize}
  \item
    some of the benefits of OO VMs don't work out in practice
  \item
    most immediate problem: it can be hard to map
    concepts of the dynamic lang to
    the host OO VM
  \item
    performance is often not improved, and can be very bad, because of the
    semantic mismatch between the dynamic language and the host VM
  \item
    poor interoperability with everything outside the OO VM
  \item
    in practice, one OO VM is not enough
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      Jython about 5 times slower than CPython
    \item
      IronPython is about as fast as CPython (but some introspection features
      missing)
    \item
      XXX something about semantic mismatch?
    \end{itemize}
  \end{block}

@@

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item
    high-level source: early design decisions not necessary
  \item
    we need a custom \emph{translation toolchain} to compile the interpreter
    to a full VM
  \item
    many aspects of the final VM are orthogonal from the interpreter source:
    they are inserted during translation
  \item
    translation aspect $\cong$ monads, with more ad-hoc control
  \end{itemize}
  \pause
  \begin{block}{
    Examples}
    \begin{itemize}
    \item
      Garbage Collection strategy
    \item
      Threading models (e.g.\ coroutines with CPS...)
    \item
      non-trivial translation aspect: auto-generating a dynamic compiler from
      the interpreter
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Links}
  \begin{block}{
    PyPy}
    \bigskip
    \hskip 1cm \url{http://codespeak.net/pypy/}
    \bigskip
  \end{block}
\end{frame}

>>>>

\title{PyPy - How to \underline{not} write Virtual Machines for \mbox{Dynamic Languages}}

\author{Armin Rigo}

@@

\date{ESUG 2007}

@@

  \begin{itemize}
  \item
    implementing dynamic languages \par(with a focus on complicated ones)
  \item
    in a context of limited resources \par(academic, open source, or
    domain-specific)
  \end{itemize}
  \pause
  \begin{block}{
    Complicated = requiring a large VM}
    \begin{itemize}
    \item
      Smalltalk (etc...): typically small core VM
    \item
      Python (etc...): the VM contains quite a lot
    \end{itemize}
  \end{block}
  \pause
  \begin{block}{
    Limited resources}
    \begin{itemize}
    \item
      Only near-complete implementations are really useful
    \item
      Minimize implementer's duplication of efforts
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Our point}
  Our point:

    \begin{itemize}
    \item
      Do not write virtual machines ``by hand''
    \item
      Instead, write interpreters in high-level languages
    \item
      Meta-programming is your friend
    \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Benefits of implementing on top of OO VMs}
  \begin{itemize}
  \item
    higher level of implementation
  \item
    the VM supplies a GC and mostly a JIT
  \item
    better interoperability than what the C level provides
  \item
    some proponents believe that eventually one single VM should be enough
  \end{itemize}
%  \pause
%  \begin{block}{
%    Python Case}
%    \begin{itemize}
%    \item
%      both Jython and IronPython integrate well with their host OO VM
%    \item
%      XXX what else?
%    \end{itemize}
%  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The problems of OO VMs}
  \begin{itemize}
  \item
    some of the benefits of OO VMs don't work out in practice
  \item
    most immediate problem: it can be hard to map
    concepts of the dynamic lang to
    the host OO VM
  \item
    performance is often not improved, and can be very bad, because of the
    semantic mismatch between the dynamic language and the host VM
  \item
    poor interoperability with everything outside the OO VM
  \item
    in practice, one OO VM is not enough
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      Jython about 5 times slower than CPython
    \item
      IronPython is about as fast as CPython (but some introspection features
      missing)
    %\item
    %  XXX something about semantic mismatch?
    \end{itemize}
  \end{block}

@@

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item
    high-level source: early design decisions not necessary
  \item
    we need a custom \emph{translation toolchain} to compile the interpreter
    to a full VM
  \item
    many aspects of the final VM are orthogonal to the interpreter source:
    they are inserted during translation
  \item
    translation aspect $\cong$ monads, with more ad-hoc control
  \end{itemize}
  \pause
  \begin{block}{
    Examples}
    \begin{itemize}
    \item
      Garbage Collection strategy
    \item
      Threading models (e.g.\ coroutines with CPS...)
    \item
      non-trivial translation aspect: auto-generating a dynamic compiler from
      the interpreter
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{For more information}
  \begin{block}{
    PyPy}
    \bigskip
    \hskip 1cm \url{http://codespeak.net/pypy/}
    \bigskip
  \end{block}
  \begin{block}{
    ``Sprints''}
    \begin{itemize}
    \item Main way we develop PyPy
    \item They are programming camps, a few days to one week long
    \item We may have one in Bern soon (PyPy+Squeak) and/or in
          Germany (JIT and other topics)
    \end{itemize}
  \end{block}
  \begin{block}{
    See also}
    Google for the full paper corresponding to these slides
    that was submitted at Dyla'2007
  \end{block}
\end{frame}
